Future challenges for Research Software Engineering

### 2025-09-25 future_challenges_for_RSEng.tex
Added some intro text to highlight the German context of the paper (#67)
* Added some intro text to highlight the German context of the paper.

* Update rebuttal notes.

* Add german to the workshop participants.

---------

Co-authored-by: Florian Goth <florian.goth@uni-wuerzburg.de>
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
was only published a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckle when they look at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs falter under algorithmic bias or unclear model provenance, it is Kim and Kay who transform the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.
